Clinical prediction with localized modeling using similarity-based cohorts: A scoping review

### 2024-06-17 docs/localized-modeling-medRxiv/localized-modeling-medRxiv.tex
prepare arxiv manuscript

\documentclass[preprint, 3p,
authoryear]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%

\usepackage[hyphens]{url}

  \journal{Artificial Intelligence In Medicine} % Sets Journal name

\usepackage{graphicx}
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
% TODO: Currently lineno needs to be loaded after amsmath because of conflict
% https://github.com/latex-lineno/lineno/issues/5
\usepackage{lineno} % add
  \linenumbers % turns line numbering on
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Clinical prediction with localized modeling using similarity-based cohorts: A scoping review},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

@@

\usepackage{lscape} \newcommand{\landscapebegin}{\begin{landscape}} \newcommand{\landscapeend}{\end{landscape}}

\begin{document}

\begin{frontmatter}

  \title{Clinical prediction with localized modeling using
similarity-based cohorts: A scoping review}
    \author[1]{Adi Cohen%
  %
  }
   \ead{ac4569@my.nsu.edu} 
    \author[2]{Patti McCall-Junkin%
  %
  }
   \ead{pattimccall.junkin@gmail.com} 
    \author[3]{Jason Cory Brunson%
  \corref{cor1}%
  }
   \ead{jason.brunson@medicine.ufl.edu} 
      \affiliation[1]{College of Osteopathic Medicine, Nova Southeastern
University, 3200 S University Drive, Davie, FL 33328}
    \affiliation[2]{Smathers Libraries Academic Research \& Consulting
Services, University of Florida, P.O. Box 117000, Gainesville, FL
32611-7000}
    \affiliation[3]{Laboratory for Systems Medicine, University of
Florida, P.O. Box 100225 JHMHC, Gainesville, FL 32610-0225}
    \cortext[cor1]{Corresponding author}

  \begin{abstract}
  Background: Applications of classical case-based reasoning (CBR) to
  clinical and health tasks have given rise to a family of techniques in
  which a statistical model is fitted to a subset of labeled cases,
  selected for relevance or similarity to an unlabeled case to which the
  model is applied. These localized models, as we propose to call them,
  may perform as well as ``black-box'' models while retaining direct
  interpretations. Their adoption has been hindered by fragmented
  development and implementation challenges. Our goal in this scoping
  review was to describe the clinical and health applications of
  localized models to date and to derive a general framework for their
  design and evaluation.

  Methods: We searched four electronic bibliographic platforms (PubMed,
  Web of Science, Academic Search Premier, and MathSciNet) during 2021
  July 19--22. We screened and reviewed entries based on four criteria
  intended to specify applications of localized models to tasks
  involving analysis of clinical or health data. Two authors divided
  title/abstract screening and collaboratively reviewed screened entries
  for inclusion. We also screened and reviewed a seed set that inspired
  the review and traced methodological citations from included entries.
  We discussed and summarized settings, tasks, and tools; identified and
  tabulated themes; and synthesized the methods into a general
  framework. We received no specific funding for this work.

  Results: Of 1,309 search results, 328 were reviewed. Of these,
  combined with 43 publications that seeded the review and 1 obtained by
  citation tracking, 25 were included in the review and methodological
  synthesis. The specificity of several search terms was poor, and
  inter-rater reliability was low. Entries were published from 1997 to
  2021, with half of these since 2015. The most common tasks were
  prognosis and diagnosis. The studies overwhelmingly focused on
  clinical data, occasionally including laboratory and image-derived
  data. Only three studies applied localized models to tasks other than
  prediction. Several studies commented on memory and runtime costs, and
  four proposed partial solutions. While few methods were reused and no
  results were reproduced, a general technique that specializes to most
  of those encountered was straightforward to describe.

  Conclusions: Localized models have potential to improve the
  performance of models used in clinical decision support tools while
  maintaining interpretability, provided computational hurdles can be
  overcome. Our search undoubtedly failed to capture all studies that
  used methods of this kind, but those we found follow a consistent
  enough design to be reproduced by a general-purpose tool. We hope that
  our review and synthesis will spur future work on efficiency,
  reproducibility, and user interactions and needs.
  \end{abstract}
    \begin{keyword}
    case-based reasoning \sep localized modeling \sep nearest
neighbors \sep clinical decision support \sep 
    scoping review
  \end{keyword}

 \end{frontmatter}

@@

A main driver of the development of clinical information systems (CIS)
and clinical decision support (CDS) tools has been to leverage the
advantages of routinely collected health data toward improving care.
These advantages include their immediate accessibility through the
institutional EHR, billing database, or other source; their specificity
to the institution and the population it serves; and the closeness in
time of the available data. Though by definition not collected for
research use, the secondary research use of routinely collected health
data has been put forth, with the monicker ``practice-based evidence'',
as an important complement to the paradigm of evidence-based practice.

Over the same period of time, interest in individualizing care from
population-derived evidence-based recommendations to specific patient
needs has driven the application of advanced computational tools,
including artificial intelligence (AI). Whereas classical models
produced formulae involving a limited set of data elements that would be
uniformly applied to all patients, AI models often process hundreds or
more variables in opaque ways to yield predictions that depend on not
only their values but their associations and their interplay with
respect to outcomes. The loss of straightforward interpretations of
these models has limited their practical uptake and motivated the
construction of explanatory statistics for opaque models as well as the
development of more interpretable complex models.

One of few methodologies to address all three of these established needs
is one we call \emph{localized modeling}, which appears to have been
introduced several times independently under different names, and in
slightly different forms. The approach is a specialized form of
case-based reasoning (CBR) that relies on a patient similarity
measure\footnote{Similarity measures between more granular units of
  analysis, such as encounters and decision points, are often still
  referred to as patient similarity measures.} to extract a cohort of
past or training cases that then inform the diagnosis, prognosis, or
care of a new or test case. While traditional CBR returns these
retrieved cases to inform human decision-making, the related nearest
neighbors (NN) technique generates predictions from cohorts
automatically via averaging (regression) or voting (classification) of
their outcomes. In contrast, localized modeling fits families of
predictive models, most commonly generalized linear models (GLMs), to
these retrieved similarity cohorts in order to generate predictions.
Localized models thus provide a way to fit fully interpretable models to
large population data in a way that is personalized to each case.

This approach harkens to the aspirational ``green button'' that would,
in response to a query, automatically retrieve patient data from an
institution's records with which to conduct on-demand retrospective
studies for an individual patient {[}1{]}. Our motivations in this
scoping review are to describe the settings in and problems with which
this approach has been tasked, to attempt to organize them within a
common framework, and to evaluate the promise they show toward achieving
this or other needs in clinical informatics. Along the way, we attempt
to reconcile terminology, summarize motivations and evaluations, and
propose valuable follow-up work.

@@

Clinical CBR emerged among rule-based approaches and other AI tools in
the development of expert systems {[}2{]}. Early implementations
realized a general workflow described as ``the four REs'', later the
``R4 cycle'' {[}2,3{]}: Given a new case (or problem), the system
\emph{retrieves} one or more past cases (solved problems) from a corpus,
\emph{reuses} these to generate an understanding of (solution to) the
new case, \emph{revises} this understanding (solution) to better fit the
new case (sometimes called ``adaptation''), and \emph{retains} the new
case and its eventual understanding (solution) in the corpus to be
retrieved and reused in future. In contrast to rule-based systems, which
are variable-based and generally interpretable as a single rule applied
to all new cases, case-based systems provide case-specific
interpretations in the form of a number of more fully understood
reference cases. While the similarity measure used in the retrieval step
need not be changed as the corpus grows, proposed measures have been
diverse, contested, and rarely systematically compared.

{[}4{]} distinguished two styles of CBR: problem-solving, which is more
procedural and used when objectives are more clearly defined, and
interpretative, which provides categorizations and justifications for
possible solutions. A similar dichotomy is commonly used to distinguish
the performance criterion for the usefulness of predictive models from
the interpretability criterion for their usability. In these terms,
localized models are highly procedural: The step of fitting a predictive
model to a retrieved cohort can be understood as an automated adaptive
strategy {[}3{]}, and the parameters that govern cohort retrieval can be
tuned alongside model hyperparameters in a ML workflow. Accordingly,
localized models have been primarily used for and evaluated on their
ability to predict classes or outcomes. Nevertheless, as some recent
applications have begun to show, they can be used to draw inferences
about the importance of different risk factors to specific individuals.
While most ML models come equipped with measures of feature importance
and model-agnostic tools can generate explanations for model
predictions, these quantifications are not directly interpretable model
components analogous to the split nodes of a decision tree (DT) or the
effect estimates of a GLM. While the patient similarity measure used to
retrieve each cohort may be complicated, the cohort itself can be
directly inspected by the user. Provided the model family fitted to the
cohorts is interpretable, the localized model as a whole inherits this
property. Thus, localized models may embody both styles of CBR.

We refer the interested reader to several previous reviews of CBR in
medicine {[}3,5,6{]}, of measures of patient similarity {[}7{]}, and of
uses of patient similarity in predictive models {[}8--10{]}. While the
reviews of CBR focus on applications using health data, the patient
similarity reviews encompass many additional types of data (various
molecular -omics, genetic tests, medical images, laboratory tests,
patient preferences, patient-reported outcomes, tracking devices, social
media) and survey a much broader scope of models (exploratory analysis
via dimension reduction and cluster analysis; risk evaluation and
outcome prediction; clinical decision support and software tools). For
the present review, we are interested in how similarity matching on
patient-level health data is used to construct local cohorts for
predictive or inferential modeling.

@@

Our goals in this review are (1) to describe applications to date of
similarity-based localized models using patient-level health data and
(2) to provide a general framework for the design and evaluation of such
localized models. Our selection of relevant papers from the search
corpus will be based on the following inclusion/exclusion criteria:

@@

We expected the general framework to come down to three choices: A
patient similarity measure, a cohort selection process, and a
statistical model family. Such a general method would be
straightforward, if laborious, to implement, and specializations of it
would account for the majority of real-world applications. A general
implementation based on this method would allow researchers to expedite
every step of the analysis process, including selection, optimization,
and evaluation, and enable sensitivity, robustness, and multiverse
analyses that help identify the most consequential choices along the
way.

@@

The structure and terms of our search strings were inspired in part by
previous reviews adjacent to our topic of interest {[}9,10{]}. We
organized the PubMed search in disjunctive normal form (an OR of ANDs).
Following the solidification of an outline, we expanded each term to
include synonyms that are similar enough to be applicable to our search.
We then evaluated the expanded search string using the PubMed Advanced
Search platform. We initially included each term and their synonyms
separately to evaluate what resulted. We pruned several terms that
returned no results (``phrases not found'') or to reduce the number of
results. At the conclusions of this process for each individual term, we
combined the search terms using disjunctive normal form.

@@

@@

Within Covidence, we screened titles and abstracts for relevance. We
decided on four eligibility criteria to expedite the screening process.
Articles must be written in English, for readability; they must be
original studies, to exclude secondary sources with duplicate
information; their use settings must be medical, clinical, or related,
to keep our review topical; and it must not be clear that their use of
our search terms was different from our intended meaning. Each of two
authors (AC and JCB) screened roughly half of the entries. They
regularly reviewed each other's decisions to improve consistency. In
cases of uncertainty, entries were included.

@@

During full text review, we decided to expand the conception of
statistical models (fourth criterion): Rather than restricting to models
that are fit and evaluated in separate steps, we chose to allow simple
statistical summaries such as mean survival {[}11{]} and survival curves
{[}12{]}. These approaches were novel to CBR and presage later
developments, so were helpful to understanding the development of
localized modeling. However, this then admitted studies that applied
conventional nearest neighbors prediction: Each retrieved cohort
consisted of the \(k\) most similar cases to the index case, and the
response for the index case was predicted to be either the mean
(continuous response) or the plurality (discrete response) of the
cohort's responses. A review of all studies that use nearest neighbors
prediction would be impractical. Because our focus is on novel
approaches that combine similarity-based retrieval and statistical
modeling of retrieved cohorts, we chose to exclude those studies whose
approach was equivalent to nearest neighbors prediction using a
conventional similarity measure.

Finally, one author (JCB) applied the same review process to the seed
set of 43 studies that inspired the review.

@@

Following the selection sources aligning with our review, we collected
characteristics of included studies (Section \ref{data-and-code}). The
features included bibliographic fields (date of publication, journal,
authors, title, keywords, DOI), study goals (objective, generalizable
knowledge, evaluation, clinical/medical domain), data sets (data source,
type of data, range of data, number of cases/incidences, number of
predictors/features), and methodological choices (types of similarity
measure, families of adaptation step/statistical model, method(s)
compared against, performance measures, results of evaluations and
comparisons, name given to modeling approach). We used these data to
detect and visualize trends amongst the included studies.

@@

Rather than an evidence synthesis characteristic of most systematic
reviews, we here pursue a methodology synthesis to harmonize largely
independent research efforts that have converged on a common modeling
technique. The goal will be to describe a unified framework for
localized models that can be used to guide future study designs and
implementations as well as more systematically evaluate variations on
the theme and measure the dependence of results on modeling choices.

@@

Figure \ref{fig:prisma} depicts our identification of studies via
databases and registers. Following the completion and input of each
search string, there were a total of 1,422 sources within all of the
platforms used. De-duplication resulted in 1,309 entries, which were
added to the title/abstract screening for review. Of these, 328 entries
met the screening criteria and were assessed through full text review.
Of these, 44 fit the original criteria, and 19 were included as distinct
from NN prediction.

@@

{\centering \includegraphics[width=0.8\linewidth]{../../fig/fig-prisma}

@@

\caption{PRISMA-S flow chart. Lowercase letters refer to items obtained from the seed set (m), the structured search (n), and citation/reference tracking (s).}\label{fig:prisma}
\end{figure}

There were 60 disagreements over inclusion versus exclusion. Inter-rater
reliability was low, at 82\% relative to a 72\% probability of random
agreement.

We then reviewed 43 studies comprising a seed set that inspired this
review. After removing duplicates and screening for eligibility, we were
left with 6 additional studies {[}12--17{]}, 1 of which was excluded
from the synthesis for using NN prediction. Finally, reference tracking
from the 43 + 328 = 371 studies assessed for eligibility led us to
identify 1 additional study that met our criteria {[}18{]}. This left us
with 19 + 1 + 5 = 25 studies included in the review and synthesis.

@@

The 25 included studies were analyzed based on several characteristics
(see Section \ref{data-and-code}), and we report and describe some
observations here (Table \ref{tab:synthesis}). The studies were
published in a variety of journals, with some of greater frequency,
though no single journal published more than 3. The journal
\emph{Evolving Systems} published 2 of the included studies,
\emph{Artificial Intelligence in Medicine} published 3, \emph{Hindawi -
Journal of Healthcare Engineering} published 3, and \emph{PLOS One}
published 2.

Another interesting pattern lay in the years of publication (Figure
\ref{fig:year}). While they trace back to the late 1990s, they have
become more common, which suggests that this is an active, though not
rapidly expanding, methodological approach. We note that, while CBR in
health and medicine originated as early as 1990, these studies did
emerge early on once CBR had established itself. So the idea has been
``in the air'' for as long as CBR has been in use.

@@

  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1215}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1602}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1215}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1823}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1547}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1050}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1547}}@{}}

@@

{[}19{]} & Prediction\hspace{6em} & Practice\hspace{6em} &
Clinical\hspace{6em} & Cross-sectional\hspace{6em} & 1,355 & 9 \\
{[}11{]} & Prognosis\hspace{6em} & Knowledge\hspace{6em} &

@@

{[}20{]} & Prediction\hspace{6em} & Practice\hspace{6em} &

@@

{[}13{]} & Diagnosis\hspace{6em} & Practice\hspace{6em} & Clinical,
Imaging\hspace{6em} & Cross-sectional\hspace{6em} & 366; 270; 560; 760 &
35; 14; 31; 8 \\
{[}21{]} & Prediction\hspace{6em} & Practice\hspace{6em} &
Clinical\hspace{6em} & Cross-sectional\hspace{6em} & 1,000; 447 & 6;

@@

{[}22{]} & Prediction\hspace{6em} & Practice\hspace{6em} &
Clinical\hspace{6em} & Cross-sectional\hspace{6em} & 2,620 & 6 \\
{[}23{]} & Prediction\hspace{6em} & Practice\hspace{6em} &

@@

{[}24{]} & Diagnosis\hspace{6em} & Knowledge\hspace{6em} &

@@

{[}18{]} & Diagnosis\hspace{6em} & Knowledge\hspace{6em} &

@@

{[}25{]} & Decision Support\hspace{6em} & Practice\hspace{6em} &
Clinical\hspace{6em} & Cross-sectional\hspace{6em} & 74 & 93 \\
{[}26{]} & Prediction\hspace{6em} & Practice\hspace{6em} &

>>>>

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}

\title{Clinical prediction with localized modeling using
similarity-based cohorts: A scoping review}

\author{
    Adi Cohen
   \\
    College of Osteopathic Medicine, Nova Southeastern University \\
  Davie, FL \\
  \texttt{\href{mailto:ac4569@mynsu.nova.edu}{\nolinkurl{ac4569@mynsu.nova.edu}}} \\
   \And
    Patti McCall-Junkin
   \\
    Smathers Libraries Academic Research \& Consulting Services,
University of Florida \\
  Gainesville, FL \\
  \texttt{\href{mailto:pattimccall.junkin@gmail.com}{\nolinkurl{pattimccall.junkin@gmail.com}}} \\
   \And
    Jason Cory Brunson
   \\
    Laboratory for Systems Medicine, University of Florida \\
  Gainesville, FL \\
  \texttt{\href{mailto:jason.brunson@medicine.ufl.edu}{\nolinkurl{jason.brunson@medicine.ufl.edu}}} \\
  }

@@

\usepackage{lscape}
\newcommand{\landscapebegin}{\begin{landscape}}
\newcommand{\landscapeend}{\end{landscape}}
\begin{document}
\maketitle

\begin{abstract}
Background: Applications of classical case-based reasoning (CBR) have
given rise to a family of techniques we call ``localized models'', in
which a statistical model is fitted to a neighborhood of labeled cases
matched by similarity to a target case. We aim to describe clinical and
health applications of localized models to date and propose a general
framework for their design and evaluation.

Methods: We searched four bibliographic platforms during 2021 July
19--22, updated 2024 January 24. We set four eligibility criteria to
identify applications of localized models to clinical and health tasks.
Two authors divided title/abstract screening and reviewed screened
entries for inclusion. We discussed settings, tasks, and tools;
identified and tabulated themes; and synthesized the methods into a
general framework.

Results: Of 1,657 search results, 360 were reviewed, then combined with
43 publications that seeded the review and 1 obtained by citation
tracking. 27 were included, published 1997--2022. The specificity of
search terms was poor, and inter-rater reliability was low. Almost all
models were predictive, the most common tasks being prognosis and
diagnosis. Most studies used clinical, occasionally laboratory and
image, data. Several addressed memory and runtime costs. A general
technique that specializes to most of those reviewed involved matching,
retrieval, fitting, and evaluation steps that could optionally be
supervised, optimized, or recursively performed.

Conclusions: Localized models have potential to improve the performance
of clinical decision support tools while maintaining interpretability,
but rigorous comparisons to competing methods must be conducted and
computational hurdles must be overcome. We hope that our review will
spur future work on efficiency, reproducibility, and user needs.
\end{abstract}

\keywords{
    case-based reasoning
   \and
    localized modeling
   \and
    nearest neighbors
   \and
    clinical decision support
   \and
    scoping review
  }

\pagebreak

@@

A main driver of the development of clinical information systems (CIS)
and clinical decision support (CDS) tools has been to leverage the
advantages of routinely collected health data toward improving care.
These advantages include their immediate accessibility through the
institutional EHR, billing database, or other source; their specificity
to the institution and the population it serves; and the closeness in
time of the available data. Though by definition not collected for
research use, the secondary research use of routinely collected health
data has been put forth, as ``practice-based evidence'', a complement to
the paradigm of evidence-based practice.

Over the same period of time, interest in individualizing care from
population-derived evidence-based recommendations to specific patient
needs has driven the application of advanced computational tools,
including artificial intelligence (AI). Whereas classical models
produced formulae involving a limited set of data elements that would be
uniformly applied to all patients, AI models often process hundreds or
more variables in opaque ways to yield predictions that depend on not
only their values but their associations and interplay with each other.
The loss of straightforward interpretations of these models has limited
their practical uptake and motivated the construction of explanatory
statistics for opaque models as well as the development of more
interpretable complex models.

One of few methodologies to address all three of these established needs
is one we call \emph{localized modeling}, which appears to have been
introduced several times independently under different names, and in
slightly different forms. The approach is a specialized form of
case-based reasoning (CBR) that relies on a patient similarity
measure\footnote{Similarity measures between more granular units of
  analysis, such as encounters and decision points, are often still
  referred to as patient similarity measures.} to extract a cohort of
past or training cases that then inform the diagnosis, prognosis, or
care of a new or test case. Traditional CBR returns these retrieved
cases to inform human decision-making; the related nearest neighbors
(NN) technique generates predictions from cohorts automatically via
averaging (regression) or voting (classification) of their outcomes. In
contrast, localized modeling fits families of predictive models, for
example generalized linear models (GLMs), to retrieved similarity
cohorts in order to generate predictions. Localized models thus provide
an individualized way to fit fully interpretable models to large
population data.

This approach harkens to the aspirational ``green button'' that would,
in response to a query, automatically retrieve patient data from an
institution's records with which to conduct on-demand retrospective
studies for an individual patient (Longhurst, Harrington, and Shah
2014). Several hurdles face the deployment of such an approach in
practice, but its benefits must first be demonstrated. Our motivations
in this scoping review are to describe the settings in and problems with
which this approach has been tasked, to attempt to organize them within
a common framework, and to evaluate the promise they show toward
achieving this or other needs in clinical informatics. Along the way, we
attempt to reconcile terminology, summarize motivations and evaluations,
and propose valuable follow-up work.

@@

Clinical CBR emerged among rule-based approaches and other AI tools in
the development of expert systems (Aamodt and Plaza 1994). Early
implementations realized a general workflow described as ``the four
REs'', later the ``R4 cycle'' (Aamodt and Plaza 1994; Begum et al.
2011): Given a new case (or problem), the system \emph{retrieves} one or
more past cases (solved problems) from a corpus, \emph{reuses} these to
generate an understanding of (solution to) the new case, \emph{revises}
this understanding (solution) to better fit the new case (sometimes
called ``adaptation''), and \emph{retains} the new case and its eventual
understanding (solution) in the corpus to be retrieved and reused in
future. In contrast to rule-based systems, which are variable-based and
generally interpretable as a single rule applied to all new cases,
case-based systems provide case-specific interpretations in the form of
a number of more fully understood reference cases. While the similarity
measure used in the retrieval step need not be changed as the corpus
grows, proposed measures have been diverse, contested, and rarely
systematically compared.

Kolodner (1992) distinguished two styles of CBR: problem-solving, which
is more procedural and used when objectives are more clearly defined,
and interpretative, which provides categorizations and justifications
for possible solutions. A similar dichotomy is commonly used to
distinguish the performance criterion for the usefulness of predictive
models from the interpretability criterion for their usability. In these
terms, localized models are highly procedural: The step of fitting a
predictive model to a retrieved cohort is an automated adaptive strategy
(Begum et al. 2011), and the parameters that govern cohort retrieval can
be tuned alongside model hyperparameters in a machine learning (ML)
workflow. Accordingly, localized models have been primarily used for and
evaluated on their ability to predict classes or outcomes. Nevertheless,
as some recent applications have shown, they can be used to draw
inferences about the importance of different risk factors to specific
individuals. While most ML models come equipped with measures of feature
importance and model-agnostic tools can generate explanations for model
predictions, these quantifications are not directly interpretable model
components analogous to the split nodes of a decision tree (DT) or the
coefficients of a GLM. While the patient similarity measure used to
retrieve each cohort may be complicated, the cohort itself can be
directly inspected by the user. Provided the model family fitted to the
cohorts is interpretable, the localized model inherits this property.
Thus, localized models may embody both styles of CBR.

We refer the interested reader to several previous reviews of CBR in
medicine (Gierl, Bull, and Schmidt 1998; Begum et al. 2011; Choudhury
and Begum 2016), of measures of patient similarity (Dai, Zhu, and Liu
2020), and of uses of patient similarity in predictive models (Welch and
Kawamoto 2013; Sharafoddini, Dubin, and Lee 2017; Parimbelli et al.
2018). While the reviews of CBR focus on applications using health data,
the patient similarity reviews encompass many additional types of data
(various molecular -omics, genetic tests, medical images, laboratory
tests, patient preferences, patient-reported outcomes, tracking devices,
social media) and survey a much broader scope of models (exploratory
analysis via dimension reduction and cluster analysis; risk evaluation
and outcome prediction; clinical decision support and software tools).
For the present review, we are interested in how similarity matching on
patient-level health data is used to construct local cohorts for
predictive or inferential modeling.

@@

Our goals in this review are (1) to describe applications of
similarity-based localized models using patient-level health data and
(2) to provide a general framework for the design and evaluation of such
localized models. We focus narrowly on localized models, rather than
broadly on patient similarity--based clinical decision support tools, so
that we may thoroughly assess their value in terms of reported
evaluations and comparisons to other methods. This also allows us to
devise a high-level framework that specializes to the majority of these
methods, which we then use to frame our discussion and recommendations.

Our selection of relevant papers from the search corpus will be based on
the following inclusion/exclusion criteria:

@@

We expected the general framework to come down to three choices: A
patient similarity measure, a cohort selection process, and a
statistical model family. A general implementation based on this method
would allow researchers to expedite every step of the analysis process,
including retrieval, optimization, and evaluation, and enable
sensitivity, robustness, and multiverse analyses that help identify the
most consequential choices along the way.

@@

The structure and terms of our search strings were inspired in part by
previous reviews adjacent to our topic of interest (Sharafoddini, Dubin,
and Lee 2017; Parimbelli et al. 2018). We organized the PubMed search in
disjunctive normal form (an OR of ANDs). Following the solidification of
an outline, we expanded each term to include synonyms that are similar
enough to be applicable to our search. We then evaluated the expanded
search string using the PubMed Advanced Search platform. We initially
included each term and their synonyms separately to evaluate what
resulted. We pruned several terms that returned no results (``phrases
not found'') or to reduce the number of results. At the conclusions of
this process for each individual term, we combined the search terms
using disjunctive normal form.

@@

We repeated the search on 2024 January 24 on the PubMed and Web of
Science databases to bring the results up to date.

@@

Within Covidence, we screened titles and abstracts for relevance. We
decided on four eligibility criteria to expedite the screening process.
Articles must be written in English, for readability; they must be
original studies, to exclude secondary sources with duplicate
information; their use settings must be medical, clinical, or related,
to keep our review topical; and it must not be clear that their use of
our search terms was different from our intended meaning. Each of two
authors (AC and JCB) screened roughly half of the entries. They
regularly reviewed each other's decisions to improve consistency. In
cases of uncertainty, entries were included. For the update, one author
(PM) screened all results for three criteria and another (JCB) screened
the survivors for the fourth criterion.

@@

During full text review, we decided to expand the conception of
statistical models (fourth criterion): Rather than restricting to models
that are fit and evaluated in separate steps, we chose to allow simple
statistical summaries such as mean survival (Mariuzzi et al. 1997) and
survival curves (Lowsky et al. 2013). These approaches were novel to CBR
and presage later developments, so were helpful to understanding the
development of localized modeling. However, this then admitted studies
that applied conventional nearest neighbors prediction: Each retrieved
cohort consisted of the \(k\) most similar cases to the index case, and
the response for the index case was predicted to be either the mean
(continuous response) or the plurality (discrete response) of the
cohort's responses. A review of all studies that use nearest neighbors
prediction would be impractical. Because our focus is on novel
approaches that combine similarity-based retrieval and statistical
modeling of retrieved cohorts, we chose to exclude those studies whose
approach was equivalent to nearest neighbors prediction using a
conventional similarity measure.

Finally, one author (JCB) applied the same review process to the seed
set of 43 studies that inspired the review. The same author later
reviewed the updated results.

@@

We next collected characteristics of included studies. The features
included bibliographic fields (date of publication, journal, authors,
title, keywords, DOI), study goals (objective, generalizable knowledge,
evaluation, clinical/medical domain), data sets (data source, type of
data, range of data, number of cases/incidences, number of
predictors/features), and methodological choices (types of similarity
measure, families of adaptation step/statistical model, method(s)
compared against, performance measures, results of evaluations and
comparisons, name given to modeling approach). We used these data to
detect and visualize trends amongst the included studies.

@@

Rather than an evidence synthesis characteristic of most systematic
reviews, we here pursue a methodology synthesis to harmonize largely
independent research efforts that have converged on a common family of
techniques. The goal will be to describe a unified framework for
localized models that can be used to guide future study designs and
implementations as well as more systematically evaluate variations on
the theme and measure the dependence of results on modeling choices.

@@

Figure \ref{fig:prisma} depicts our identification of studies via
databases and registers. Following the completion and input of each
search string, there were a total of 1,817 sources within all of the
platforms used. De-duplication resulted in 1,657 entries, which were
added to the title/abstract screening for review. Of these, 360 entries
met the screening criteria and were assessed through full text review.
Of these, 51 fit the original criteria, and 21 were included as distinct
from NN prediction.

@@

{\centering \includegraphics[width=0.8\linewidth]{Fig1}

@@

\caption{PRISMA-S flow chart. Lowercase letters refer to items obtained from the seed set (m), the structured search (n), and citation/reference tracking (s). Within the structured search results, summands correspond to original and update.}\label{fig:prisma}
\end{figure}

From the first search results, there were 60 disagreements over
inclusion versus exclusion. Inter-rater reliability was low, at 82\%
relative to a 72\% probability of random agreement. Only one author was
available to review full texts from the update search.

We then reviewed 43 studies comprising a seed set that inspired this
review. After removing duplicates and screening for eligibility, we were
left with 6 additional studies (Park, Kim, and Chun 2006; Lowsky et al.
2013; Lee, Maslove, and Dubin 2015; Ng et al. 2015; Lee 2017; N. Wang et
al. 2019), 1 of which was excluded from the synthesis for using NN
prediction. Reference tracking from the 43 + 360 = 403 studies assessed
for eligibility led us to identify 1 additional study that met our
criteria (Kasabov and Hu 2010). This left us with 21 + 1 + 5 = 27
studies included in the review and synthesis.

@@

The 27 included studies were analyzed based on several characteristics,
and we report and describe some observations here (Table
\ref{tab:synthesis}). The studies were published in a variety of
journals, with some of greater frequency, though no single journal
published more than 3. The journal \emph{Evolving Systems} published 2
of the included studies, \emph{Artificial Intelligence in Medicine}
published 3, \emph{Hindawi - Journal of Healthcare Engineering}
published 3, \emph{Evolving Systems} published 2, and \emph{PLOS One}
published 2.

Another interesting pattern lay in the years of publication (Figure
\ref{fig:year}). While these studies trace back to the late 1990s, they
have become more common, which suggests that this is an active, though
not rapidly expanding, approach. We note that CBR in health and medicine
originated as early as 1990, and localized modeling emerged soon after
CBR had established itself; the idea has been ``in the air'' for as long
as CBR has been in use.

\small

@@

  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1148}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1585}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1202}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1967}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1530}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1038}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1530}}@{}}

@@

Yearwood and Wilkinson (1997) & Prediction\hspace{6em} &
Practice\hspace{6em} & Clinical\hspace{6em} &
Cross-sectional\hspace{6em} & 1,355 & 9 \\
Mariuzzi et al. (1997) & Prognosis\hspace{6em} & Knowledge\hspace{6em} &

@@

Wyns et al. (2004) & Prediction\hspace{6em} & Practice\hspace{6em} &

@@

Park, Kim, and Chun (2006) & Diagnosis\hspace{6em} &
Practice\hspace{6em} & Clinical, Imaging\hspace{6em} &
Cross-sectional\hspace{6em} & 366; 270; 560; 760 & 35; 14; 31; 8 \\
Song and Kasabov (2006) & Prediction\hspace{6em} & Practice\hspace{6em}
& Clinical\hspace{6em} & Cross-sectional\hspace{6em} & 1,000; 447 & 6;

@@

Elter, Schulz-Wendtland, and Wittenberg (2007) & Prediction\hspace{6em}
& Practice\hspace{6em} & Clinical\hspace{6em} &
Cross-sectional\hspace{6em} & 2,620 & 6 \\
Xu et al. (2008) & Prediction\hspace{6em} & Practice\hspace{6em} &

@@

López et al. (2011) & Diagnosis\hspace{6em} & Knowledge\hspace{6em} &

@@

Kasabov and Hu (2010) & Diagnosis\hspace{6em} & Knowledge\hspace{6em} &

@@

Verma et al. (2015) & Decision Support\hspace{6em} &
Practice\hspace{6em} & Clinical\hspace{6em} &
Cross-sectional\hspace{6em} & 74 & 93 \\
Liang, Hu, and Kasabov (2015) & Prediction\hspace{6em} &
